Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-14 pkg/circtree/inst/wind_paper/draft/draft.Rnw
some touch-ups in Sections 4 and 5
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2933 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e596d25751143dde1f4ed92025bbd31262fd447e

% -------------------------------------------------------------------
% Case study
% -------------------------------------------------------------------
\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
As motivated in Section~\ref{sec:introduction}, accurate forecasts of wind
directions are of great importance for risk management in various fields such
as agriculture, energy production or aviation. For example, in order to direct
airplanes to a safe landing, precise knowledge of wind direction for the next
hour(s) at the respective airport is highly desirable and adequate prediction
methods are required. This section exemplifies the use of circular regression
trees and forests with wind direction forecasts for two Austrian airports --
one in flat terrain, the other one in mountainous terrain. The results are
benchmarked against alternative probabilistic forecasting methods. The study
is based on $1$\,h and $3$\,h forecasts employing lagged observations in the
vicinity of the airports as possible predictor variables.

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airports (left) and Vienna International Airport (right). 
    For Innsbruck, $4$~stations at the airport and $6$~stations along the
    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
    and $13$~stations within the vicinity of approximately 30\,km are used.
    Elevation data are obtained from the TandDEM-X digital elevation model with a 
    horizontal resolution of $90$~m \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min-average
of wind direction measurements at Innsbruck Airport (INN) and Vienna International Airport (VIE) on
an hourly temporal resolution. As predictor variables, we use temporal
information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities such as wind direction, wind speed, temperature, air
pressure and humidity, all lagged by one or three hours according to the
respective forecasting step. The meteorological variables are measured either
directly at the airports or within their vicinities. For Innsbruck,
measurements at the airport and along the intersecting valleys are used,
whereas, for Vienna, measurements at the airport and within its vicinity of
approximately 30\,km are used. A topographical overview of the airports and
their surrounding areas with the station sites employed in this study is
provided in Figure~\ref{fig:overview}. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

The data used in this study consists of five subsequent years from January 2015
to December 2018. After first eliminating predictor variables with more than
5\% missing values and then points with missing observations, the data set
consists of 41\,979 dates and 260 covariates for Innsbruck, and of
38\,985 dates and 494 covariates for Vienna, respectively.

@@

\subsection{Models and evaluation}
\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches used in the meteorological field, while the third is a
state-of-the-art GLM to forecast circular response variables. The next
paragraphs introduce the reference methods and provide the corresponding model
specifications used in this study for all considered models including the
circular regression tree and forest.

@@

    wide range of applications and climatologies are often also used as a baseline for the
    validation of newly developed forecasting systems
    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. Usually, the performance of
    climatological models strongly depends on the temporal dependency of the
    quantity itself, either on a daily or yearly basis. Wind direction at a single
    location is often quite unstable and only shows a daily cycle or an annual seasonality
    in case of predominant local wind systems like the ones at Innsbruck introduced
    in the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model
    in this study provides a very simple implementation of a baseline model, which
    might experience serious performance losses in case of random temporal
    variabilities or abrupt changes in the employed response variable wind
    direction.

@@

    Instead of simply fitting a climatological model over all dates, separate
    models are estimated rolling over all observation dates, using the same time of
    $31$~days centered around the day of interest over all available years in the
    training data.  To gain full probabilistic climatological forecasts, we employ
    the steps described in Section~\ref{sec:prob_circ} and fit the distribution
    parameters of the von Mises distribution via maximum likelihood estimation.
    This approach follows \citet{Vogel+Knippertz+Fink:2018} and is discussed in a
    comprehensive summary on different time-adaptive training schemes in

@@

    in a time series. Hence, it is another objective measure in the verification of
    weather forecasts and also often applied as a reference model
    \citep{noaasnationalweatherservice:2019}. For short forecast steps under stable
    atmospheric conditions, persistence can be a very good estimate, but as soon as
    the regime changes during the forecast period, forecasts based on persistence
    may also fail remarkably. Therefore, whether the persistence can provide
    overall reliable information, heavily depends on the variability of the
    variable of interest and the length of the forecasting step.

@@

    classes (argument \code{nmax = c("yx"\,=\,Inf, "z"\,=\,50)}). Contrary to a

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for the airports Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation
years, yielding yearly mean values for every hour per calendar month, with a
lower score indicating better performance. The circular forest overall provides
the best predictive performance, followed by the circular regression tree and
the persistency model for both stations at both forecast steps; except for the
$3$\,h~forecasts at Innsbruck where the persistency model is outperformed by
all others. In comparison to the circular regression tree and forest, for both
stations and forecast steps, the climatological model and the linear model show
clearly higher CRPS values and hence a lower predictive performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores of wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
    airports Innsbruck and Vienna. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    regression tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains only a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift, e.g., from up-valley to
down-valley. Due to the few wind regimes the rather inflexible climatological
and linear models score relatively well with similar CRPS values as the other
models (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a change of the wind regime seem to
lead to a higher variation in the predictive performance for all models in
comparison to Vienna; this variation is especially high for the persistency
model due to its strong vulnerability to abrupt wind shifts. On the contrary,
at Vienna the geographic surroundings do not constrain any wind direction which
leads to smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and
forests are established by coupling model-based recursive partitioning with the
von Mises distribution. By separating the data into more homogeneous
subgroups, possible difficulties in circular regression are bypassed as
covariates are solely considered for splitting and group-specific models are
fitted as intercept-only models. In addition, by specifying the von Mises
distribution for each node and allowing for splits in both distribution
parameters $\mu$ and $\kappa$, fully probabilistic forecasts are provided.

The performance of the novel circular trees and forests is assessed in a 
case study for shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchline models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM are evaluated based on proper scoring rules. In summary, the
circular trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods. Another possible
improvement for obtaining more parsimonious circular trees is to consider
splitting circular covariates into two circle segments by searching two split
points simultaneously rather than sequentially at different depths. While this
might slightly improve the predictive performance of circular trees, this
should not affect the performance of circular forests, as these consist of
very large trees with many different splits.

>>>>

% -------------------------------------------------------------------
% Case study
% -------------------------------------------------------------------
\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
As motivated in Section~\ref{sec:introduction}, accurate forecasts of wind
directions are of great importance for risk management in various fields such
as agriculture, energy production, or aviation. For example, in order to direct
airplanes to a safe landing, precise knowledge of wind direction for the next
hour(s) at the respective airport is highly desirable and adequate prediction
methods are required. This section exemplifies the use of circular regression
trees and forests with wind direction forecasts for two Austrian airports --
one in flat terrain, the other one in mountainous terrain. The results are
benchmarked against alternative probabilistic forecasting methods. The study
is based on $1$\,h and $3$\,h forecasts employing lagged observations in the
vicinity of the airports as possible predictor variables.

\subsection{Data}
%%
\begin{figure}[t]
  \setkeys{Gin}{width=1.0\textwidth}
  \centering
  \includegraphics{_plot_circforest_stationmap_combined.pdf}
  \caption{Overview of the study area for Innsbruck Airports (left) and Vienna International Airport (right). 
    For Innsbruck, $4$~stations at the airport and $6$~stations along the
    intersecting valleys are used, whereas, for Vienna, $9$~stations at the airport
    and $13$~stations within the vicinity of approximately 30\,km are used.
    Elevation data are obtained from the TandDEM-X digital elevation model with a 
    horizontal resolution of $90$~m \citep{Wessel:2018}.}
  \label{fig:overview} 
\end{figure}
%%
\label{sec:wind:data}
The circular response variable considered in this case study is a $10$\,min-average
of wind direction measurements at Innsbruck Airport (INN) and Vienna International Airport (VIE) on
an hourly temporal resolution. Temporal information and $1$-hourly resolved $10$\,min mean observations of various
meteorological quantities are used as predictor variables, including wind direction, wind speed, temperature, air
pressure and humidity, all lagged by one or three hours according to the
respective forecasting step. The meteorological variables are measured either
directly at the airports or within their vicinities. For Innsbruck,
measurements at the airport and along the intersecting valleys are used,
whereas, for Vienna, measurements at the airport and within its vicinity of
approximately 30\,km are used. Figure~\ref{fig:overview} provides a topographical
overview of the airports and their surrounding areas with the station sites employed
in this study. In addition, we use derived quantities
such as 3-hourly means, minima and maxima, as well as 1-and 3-hourly temporal
changes and spatial differences towards the airport of the respective
quantities. An overview of the employed data sets can be found in
Table~\ref{tab:data}.

The data used in this study consists of four subsequent years from January 2015
to December 2018. After first eliminating predictor variables with more than
5\% missing values and then time points with any missing observations, the data set
consists of 41\,979 time points and 260 covariates for Innsbruck, and of
38\,985 time points and 494 covariates for Vienna, respectively.

@@

\subsection{Models and evaluation}
\label{sec:wind:models}
For a fair evaluation of circular regression trees and forests, and to
investigate whether they can be applied as a reasonable alternative to already
existing approaches, three additional statistical models are employed in this
study for probabilistic forecasting of wind directions. Two of them are based
on existing approaches used in the meteorological field, while the third is a
state-of-the-art GLM-type model to forecast circular response variables.

@@

    wide range of applications. While forecasts based on climatologies, by
    construction, do not adapt to the current weather situation they are still a
    useful baseline for the validation of newly developed forecasting systems
    \citep{Simon+Umlauf+Zeileis:2017, Stauffer+Mayr+Messner:2017}. 

@@

    Specifically, the climatology employed in the following uses all observations
    at the same time (to adapt to daily cycles) in a window of $31$~days centered
    around the day of interest (to adapt to seasonal cycles)
    in all available years in the sample. Based on these observations a probabilistic
    model is obtained by maximum likelihood estimation as described in
    Section~\ref{sec:prob_circ}. This approach follows \citet{Vogel+Knippertz+Fink:2018}
    and is discussed in a comprehensive summary on different time-adaptive training schemes in

@@

    in a time series. Like the climatology it is a very basic prediction model that
    is often applied as a baseline reference in weather forecasts
    \citep{noaasnationalweatherservice:2019}. Especially in nowcasting tasks with very short
    forecasting steps, the persistence can provide very good estimates.

@@

    classes (argument \code{nmax = c(yx = Inf, z = 50)}). Contrary to a

@@

\subsection{Results} 
This section provides a detailed analysis on the predictive performance of the
different proposed statistical models applied to probabilistic wind direction
forecasting. To ensure a comprehensive comparison of the models, wind direction
forecasts are evaluated for two different lead times at two airports with
different climatological site characteristics.
Figure~\ref{fig:boxplot_crpsraw} shows the CRPS values of the employed models
at forecast steps 1\,h and 3\,h for Innsbruck (Panels~a,\,c) and
Vienna (Panels~b,\,d). The scores are aggregated over the five validation
years, yielding yearly mean values for every hour per calendar month, with a
lower score indicating better performance. The circular forest overall provides
the best predictive performance, followed by the circular regression tree and
the persistency model for both stations at both forecast steps; except for the
$3$\,h~forecasts at Innsbruck where the persistency model is outperformed by
all others. In comparison to the circular regression tree and forest, for both
stations and forecast steps, the climatological model and the linear model show
clearly higher CRPS values and hence a lower predictive performance.
%%
\begin{figure}[t]
  \centering
  \setkeys{Gin}{width=1.0\textwidth}
  \includegraphics{_plot_circforest_validation_crpsraw_agg_comparison_with_lowff_v14b.pdf}
  \caption{CRPS skill scores of wind direction forecasts based on the full
    predictive von~Mises distribution for $+1$\,h and $+3$\,h forecasts at the
    Innsbruck Airport and Vienna International Airport. Each box-and-whisker contains $24$~hourly scores
    for each of the $12$~months averaged over the $5$~validation years which
    yields a total of $288$~yearly mean values. The scores are shown for the
    climatological, persistency and the linear model as well as for the circular
    regression tree and forest.}
  \label{fig:boxplot_crpsraw} 
\end{figure}
%%
The different site characteristics of the airports Innsbruck
(Figure~\ref{fig:boxplot_crpsraw}\,a,\,c) and Vienna
(Figure~\ref{fig:boxplot_crpsraw}\,b,\,d) seem to have an effect on the
absolute level of the model performances and on their respective predictive
performance variances. At Innsbruck, due to the surrounding mountains only a
limited number of possible wind directions exists, namely the three wind
regimes discussed for Figure~\ref{fig:tree_ibk} in
Section~\ref{sec:introduction}. Therefore, for Innsbruck the wind direction
remains rather constant in one of these possible states, but once a change
takes place it is mostly a major wind direction shift, e.g., from up-valley to
down-valley. Due to the few wind regimes the rather inflexible climatological
and linear models score relatively well with similar CRPS values as the other
models (Figure~\ref{fig:boxplot_crpsraw}\,a,\,c). In addition, at Innsbruck the
potential high prediction errors in case of a change of the wind regime seem to
lead to a higher variation in the predictive performance for all models in
comparison to Vienna; this variation is especially high for the persistency
model due to its strong vulnerability to abrupt wind shifts. On the contrary,
at Vienna the geographic surroundings do not constrain any wind direction which
leads to smaller and less abrupt changes in the wind direction as well as less
pronounced wind regimes. This seems to weaken the predictive performance of the
climatological and linear models, and to reduce the performance variability for
all models (Figure~\ref{fig:boxplot_crpsraw}\,b,\,d).

@@

% -------------------------------------------------------------------
% Summary
% -------------------------------------------------------------------
\section{Summary and conclusion}
\label{sec:summary}
Extending the toolbox for modeling circular data, circular regression trees and
forests are established by coupling model-based recursive partitioning with the
von Mises distribution. By separating the data into more homogeneous
subgroups, possible difficulties in circular regression are bypassed as
covariates are solely considered for splitting and group-specific models are
fitted without further covariates. In addition, by specifying the von Mises
distribution for each node and allowing for splits in both distribution
parameters $\mu$ and $\kappa$, fully probabilistic forecasts are provided.

The performance of the novel circular trees and forests is assessed in a 
case study for shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchline models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In summary, the
circular trees and forests have the highest predictive performance in this
setting. For cases without changes in the wind regime, lagged response values
provide already highly skillful estimates leading to a good performance of the
persistency model as observed for shortterm wind direction forecasts in this
study. While in these cases the trees and forests also benefit from the highly
informative lagged response, they can compensate for a lower information of
this covariate by incorporating other quantities and possible interactions of
these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods. Another possible
improvement for obtaining more parsimonious circular trees is to consider
splitting circular covariates into two circle segments by searching two split
points simultaneously rather than sequentially at different depths. While this
might slightly improve the predictive performance of circular trees, this
should not affect the performance of circular forests, as these consist of
very large trees with many different splits.
